Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems

### 2019-10-19 repro/papers/mps_theory/response.tex

\title{Responses to reviewer's comments on ``Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems''}
\date{}
\address{
        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
        77005-1892 USA
}
\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}

\righthead{Responses}

\maketitle
\parskip 12pt
\section{Introduction}
We thank the reviewers for insightful and constructive comments. Their insights led to many changes in this manuscript. We trust that the revised manuscript may answer the questions posed, and generally provide a much improved overview of our work.

\section{Reviewer 1}

Comment: The article presents an extension to the theory on moment conditions for finite differences to multipole point sources. The moment conditions are first derived in a continuous setting followed by the derivation of a discrete set of moment conditions. Further, a convergence theory for staggered grid finite-difference schemes on a family of first order systems of hyperbolic PDEs with multipole source terms is presented. The authors conclude the article by testing the convergence by numerical experiments on the acoustic wave equation.

In the introduction the authors claim that the presented work is an extension of the theory on point source discretization in finite differences. The moment conditions in the first section appears as a valuable extension to current theory, but the convergence analysis is restricted to a specific formulation of hyperbolic PDEs solved with staggered grid finite differences. My interpretation is that these results are not general for all stable finite difference discretisations. If the theory is more general than my interpretation, this must be stated more clearly. Otherwise, the reason for the restriction to staggered grid-finite differences should be explained.

Response: The introduction has been rewritten to emphasize the focus on staggered grid discretizations for symmetric hyperbolic systems of ``div-grad'' form.  The latter does not seem a severe constraint to us, as several physically important wave theories can be cast in that form, and others (e.g. several forms of viscoelasticity) can be viewed lower-order perturbations of systems of this type. The focus on staggered grid schemes is a more serious limitation. It is conceivable that results like those proven here could be established for other stable finite difference schemes, or for finite element schemes of various types, however such generalization is beyond the scope of this paper. The key ingredient in our arguments is the discrete energy estimate proved in the theory section. However as is well-known such an estimate is only a sufficient condition for stability, hence convergence.

%I think the for the most part, the paper does a good job in specifying that the analysis and results are limited to staggered-grid FD schemes. Hopefully this is more apparent now that the theory section has been cleaned up. I added some extra sentences in the conclusion section to emphasize the integral components of the theory. The motivation for using staggered-grid methods was briefly mentioned in the intro (page 4, bottom paragraph). Not sure if it is worth mentioning again. Would it be a good idea to “staggered-grid” somewhere in the title?

Comment: In segments of the article, the line of argument is difficult to follow. For example, the subsection "Examples" on p. 23 appears out of context and the subsection is not well written. In the introduction of this subsection numerical convergence tests in the "next section" are mentioned but the next section does not contain such tests. Figure references are also missing in this section. This subsection must be rewritten and perhaps moved to the section "numerical tests and results".

Response: This section has been cleaned up, and figure references added.

Comment: The structure of the convergence theory is difficult to follow since the more general system which is actually analyzed (21) is mixed with an example (23). I think the example causes confusion about what is actually proven. I believe that the example would fit better in the section "numerical tests and results”.

Response: The convergence theory section has been extensively reworked. All references to the acoustic case have been moved from the theory section to a subsection titled “Acoustic Case” at the end of the respective section.

Comment: The way of presenting the convergence result is non-standard. I suggest that you perform a convergence study with several grid refinements presented similar to figure 7 in Petterson et al. 2016. If if this is not added, a motivation of the current form of convergence test is needed.

Response: Convergence rates are presented in this manner to demonstrate their spatial dependence, in particular, relative to source location in 2D space. For example, this presentation highlights the degradation in convergence rate far from the source location when using source approximations with the wrong approximation order. The convergence rates reported here are equivalent to the slopes of typical convergence figures in log-log plots.

Comment: In figure 6 (a) a green line which corresponds to third order convergence appears vertically in the figure. In figure 7 (a) one can also detect some green areas surrounding the source. Can you comment on this locally lower convergence rate?

Response: I think the author mean figures 6(c) and 7(c). The green line in fig 6(c) coincides with the nodal plane of the dipole.

\section{Reviewer 3:}

Comment: Your analysis is restricted to the Cauchy problem (periodic case), but that is not stated anywhere (please state).  Many applications require treatment of point sources on the boundary. In this case, the moment conditions need to be modified to account for the quadrature rule used by the difference scheme, as shown in Petersson 2016.

Response: Quite correct, and intentional. We do not limit the analysis to the periodic case, but to the full-space case, assuming always that solutions have compact support in space for each time.  It is certainly possible to extend the results to some combinations of boundary conditions and interior and boundary schemes, so long as the latter admit energy estimates. We know of no such staggered grid boundary schemes for the elastic free surface of order higher than two, for example (though we are aware of widely used approximations, such as Robertsson's vacuum scheme). Also we do not study sources supported on the boundary. All of these other questions are interesting, but beyond the scope of the paper. We include explicit statements to that effect in the rewritten introduction.

%Boundary conditions are not required in the analysis, except implicitly in satisfying the skew-adjoint relations in the continuum and discrete systems. In the numerical examples, we use PML boundaries to simulate an unbounded domain, though we implement reflective boundary conditions at the computational domain, consistent with the theory. Source discretization near boundaries is not considered here. Not sure what could be added to the paper.

Comment: Before (16) on page 19, invervals $\rightarrow$ intervals

Response: fixed

Comment: On page 33 "corollary above", what corollary? do you mean theorem?

Response: fixed

Comment: Page 36, did you mean $\tilde{u}$ and not $\tilde{p}$?

Response: fixed.

Comment: Page 37, does $\pm$ indicate addition and subtraction?

Response: Yes.

Comment:  You rely on $P_h = -P_h^T$ for your proofs, but this property is never stated (only for the continuous problem).  It is now mentioned explicitly in the presentation of the general continuum problem and its discretization by the skew-adjoint relations (23) and (25) respectively.

Response: Actually this statement (that $P_h$ must be skew adjoint) was an error on our part, and has been removed entirely from the rewrite - the acoustic example, for instance, does not have this property. The matrix operator $\left(\begin{array}{cc}0 & P^T\\-P & 0\end{array}\right)$ is skew adjoint, for any P of the type considered here (continuous or discrete).

Comment: Page 38, How can you bound ${\cal E}_2$ using $L^2$-error estimates of finite difference solutions for smooth problems, when the solution to the problem is not smooth?

Response: $(\tilde{u}^{n-1/2},\tilde{v}^n)$ solves a FD approximation to the adjoint problem with smooth data $\tilde{f},\tilde{g}$, hence satisfies an energy estimate. The point of using the adjoint formulation is that it replaces inner products of  singular solutions and smooth test functions with inner products of singular sources and smooth solutions with the test functions serving as sources.

Comment: At no point in your proof do you make use of the fact that you have a staggered finite difference discretization.

Response: This is an astonishing statement, in view of the formulation of the abstract staggered grid scheme (equation 24) and its use througout the rest of the paper. We are at a loss as to what the reviewer means.

Comment: However, if the discretization is replaced by the collocated grid, central finite discretizations (e.g., -1/2h, 0,1/2h), then smoothness conditions are necessary for convergence. To prove convergence without imposing smoothness conditions, I believe, at some point you have to invoke the property that the Nyquist mode $(-1)^j$ does not lie in the nullspace of the staggered grid difference operators.

Response: if the Nyquist mode lay in the nullspace of the finite difference propagator, then an energy estimate of the type we prove would be impossible. Conversely, the energy estimate shows that no such phenomenon can occur, provided that the CFL criterion is satisfied.

%We can apply L_2-error estimates to \mathbcal{E}_2 since \delta{\tilde u} and \delta{\tilde v} refer to the finite difference error for a problem with smooth sources (\tilde f, and \tilde g), i.e., see equation (35).  We did most of the heavy lifting for the proof of theorem 8 (weak convergence) a bit ahead of the actual theorem.  May have contributed to some confusion for the reviewer?

Comment: Page 42, how do you define the minimum number of grid points per wavelength? Usually, the frequency at 5\% of
peak amplitude in the spectral domain is defined as $f_{max}$ and then the minimum wavelength is estimated by $ \lambda_{min} = c_{min}/f_{max}$, where $c_{min}$ is the minimum wavespeed. For Ricker wavelets, one obtains $f_{max} = 2.5 f_p$ where $f_p$ is the peak frequency.

%Dr. Symes, not sure how to handle this? Would it be sufficient to reference a paper stating the rule of thumb of 10 grid points per wavelength for 2-2 and 5 grid points per wavelength for 2-4? If so, could you remind me of the reference?

Response: We have removed this paragraph. In our study of convergence rates, gridpoints per wavelength is not a relevant concept, though it might be used to encapsulate the results of such a study.

Comment: Page 43, You state "... our narrow source approximations seem twice as wide and therefore smoother, from the point of view of the difference operator $P_h$. For this reason, smoothness conditions are not necessary for staggeredgrid finite difference schemes". I do not follow this argument. Why would stencil width imply smoothness? As previously mentioned, smoothness conditions are not necessary for convergence when the Nyquist mode does not lie in the nullspace of the difference operator.

%Response: The stencil width and smoothness comment was referring to smoothness in Fourier space. For example, the Fourier transform of a wider the hat function is a narrower $sinc^2$ function, thus the contribution from spurious modes is diminished. Incidentally, source discretizations presented by Petersson 2016 that satisfy m-moment and ssmoothness conditions (with m=s) coincide with approximations that satisfy m-moment conditions and have two times the stencil width relative to the finite difference operator. These are the same source approximations used here.

Response: we have removed this discussion, as it is redundant. We have shown in the theory section that for the class of schemes considered here, the moment conditions imply convergence for multipole rhs (in the weak sense, of course), without need to appeal to any additional conditions such as the smoothness condition of  Petersson et al.

Comment: Page 44, Instead of saying convergence theory "applicable to a large set of wave propagation problems", why not
simply say "wave equations"? In other places you say "symmetric hyperbolic systems, which also can change to
"wave equations" to be more specific.

%Dr. Symes, any thoughts on this?

Response: We are not sure why the referee makes these suggestions. Symmetric hyperbolic system are first order systems of PDEs with wave-like solutions. They do not encompass all wave equations (or even the ``classic'' wave equation), but most wave propagation dynamics can be (re)formulated in terms of such systems. So as we say we are mystified, and with apologies for being so obtuse, have left the prose in question intact.

### 2019-10-19 repro/papers/mps_theory/paper_orig.tex
repro docs dir added

%\title{Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems}
%\date{}
%\address{
%        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
%        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
%        77005-1892 USA
%}
%\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}

%\righthead{Multipole Source Discretizations}

%\maketitle
%\parskip 12pt

@@

Seismic sources are commonly idealized as point-sources due
to their small spatial extent relative to seismic wavelengths.
The acoustic isotropic point-radiator is inadequate as a model 
of seismic wave generation for seismic sources that are known to exhibit directivity. 
Therefore, accurate modeling of seismic wavefields must include source 
representations generating anisotropic radiation patterns. 
Such seismic sources can be modeled as 
{\em multipoles}, that is, time-dependent linear combination of
spatial derivatives of the spatial delta function. Since the solutions of linear hyperbolic
systems with point-source right hand sides are necessarily singular, standard results on
convergence of grid-based numerical methods (finite difference or
finite element) do not imply convergence of numerical solutions.
We present a method for discretizing multipole sources 
in a finite difference setting, an extension of the moment matching 
conditions developed for the Dirac delta function in other applications.
along with numerical evidence demonstrating the accuracy of these approximations.
Using this analysis, we develop a weak convergence theory for the discretization of 
a family of symmetric hyperbolic systems of first-order partial
differential equations, with singular source terms, solved via
staggered-grid finite difference methods: we show that
grid-independent space-time averages of the numerical solutions
converge to the same averages of the continuum solution, and provide
an estimate for the error in terms of moment matching and truncation
error conditions.
Numerical experiments confirm this result, but also suggest a stronger
one: optimal convergence rates appear to be achieved
achieved point-wise in space away from the source.

@@

%Seismic sources are commonly idealized as concentrated at a source point due
%to their small spatial extent relative to seismic wavelengths.
Linear wave propagation problems modeling physical systems with
concentrated (in space) energy sources arise in a number of
applications, such as seismology, nondestructive material evaluation,
and radar. A familiar example of a wave propagation model with
spatially concentrated 
%energy
energy
source is the isotropic point-radiator problem
for the acoustic wave equation in Euclidean 3-space \cite[]{CourHil:62}:
\begin{equation}\label{eq:isorad}
\begin{split}
       \frac{1}{c^2} \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - \nabla^2
  p(\mathbf x,t) & = f(\mathbf x,t) \equiv w(t)\delta(\mathbf x),\\ 
p(\mathbf x,t) & =  0,  \quad t<<0,
\end{split}
\end{equation}
$\delta(\mathbf x)$ being the Dirac delta function.
The solution $p$ is spherically symmetric:
\begin{equation}
\label{eq:green3d}
	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi r}, \quad 
	r= \sqrt{\mathbf x^T \mathbf x}.
\end{equation}

% here is a heaven-sent opportunity to tell the reader what the paper
% is about

The form \ref{eq:green3d} of the isotropic point radiator solution
suggests an obstacle to successful  approximation by gridded (finite
difference, finite element,...) numerical methods: the fundamental
rationale for the accuracy of such methods, namely truncation error
analysis based on Taylor expansion, is invalid in any neighborhood of
the source point $\mathbf x = \mathbf 0$, as the solution is singular
there. Moreover, due to the wave nature of both continuum and gridded
approximate solutions for problems like \ref{eq:isorad}, errors
arising near the source point propagate throughout the region in which
the solution is computed. Examples presented later in this paper
illustrate this generation and propagation of error due to source singularity.

The main result of this paper is that certain special approximations
to wave dynamics problems such as \ref{eq:isorad} with (singular)
right-hand sides of point support lead to finite difference solutions
that converge globally in a weak sense - that is, averages over fixed
regions converge as the grid cell size is reduced.

In the remainder of this introduction, we describe the class of wave
dynamics problems to which our arguments apply, the representation and approximation
of right-hand sides,  the type of numerical
approximation used in our work, and the weak convergence of approximate solutions.

We focus our attention on a symmetric hyperbolic systems of the form
\begin{equation}\label{eq:hyper_fam}
\begin{split}
	A \frac{\partial}{\partial t}u + P^T v &= f \\
	B \frac{\partial}{\partial t}v  - P u & = g
\end{split}
\end{equation}
connecting dynamical fields $u,v$, energy source fields $f,g$, and symmetric
positive definite matrix fields $A,B$. The operator $P$ is a
constant-coefficient differential operator of order 1; $P^T$ denotes its
formal adjoint or transpose.
Many important physical systems supporting wave propagation can be
cast in the form \ref{eq:hyper_fam}. These include linear
elastodynamics and Maxwell's equations. In all such examples, the
matrices $A$ and $B$ encode time-independent physical parameters characteristic of the
continuum mechanical model. For example, elastodynamics may be cast in
the form \ref{eq:hyper_fam}, with $A$ being the compliance tensor,
$B$ being the material density multiplied by the $3 \times 3$
identity, and $P$ is the strain operator. As is well-known, the acoustic
wave equation \ref{eq:isorad} is also equivalent to such a system, in the
sense that its solution $p$ can be identified with $u$ above, with $v$
chosen as the indefinite time integral of $-\nabla p$. In this case,
$A$ is the $1 \times 1$ matrix with entry $c^2$, $B$ is the $3
\times 3$ identity matrix, and $P$ is the gradient operator. The resulting system expresses
acoustodynamics with material density normalized to 1, and $c^2$
identified with bulk modulus of the fluid material. The dynamic fields
$p$ and $v$ represent pressure and particle velocity respectively.

Sources of small support relative to typical wavelengths are reasonable models
for some seismic energy generation mechanisms. Such sources may be
approximated by point support right hand sides such as occur in the
isotropic radiator problem \ref{eq:isorad}. Of course, ``functions''
of space and time having point support in space are not really
functions at all, but (Schwarz) distributions. 
Peetre's Theorem \cite[]{Horm:69} implies that any distribution in
${\bf R}^{d+1}$ 
supported on $\{\mathbf x^* \} \times {\bf R}$, ${\bf x^*} \in {\bf R}^d$, takes the form of a finite series
\begin{equation}\label{eq:MPSappx}
        f(\mathbf x,t) = \sum_{|\mathbf s|=0}^N w_{\mathbf s}(t) \; D^{\mathbf s}\delta(\mathbf x-\mathbf x^*),
\end{equation}
in which we have introduced {\em multi-index notation}: for spatial
dimension $d$ and multi-index (integer d-tuple) ${\bf s} =
(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative operator, denoted $D^{\bf s}$,
and its (total) order $|{\bf s}|$, are defined as
\begin{equation}\label{eq:PDO}
        D^{\mathbf s} = \prod_{i=1}^d \left( \frac{\partial}{\partial x_i}\right)^{s_i}, \quad |\mathbf s| = \sum_{i=1}^{d} s_i.
\end{equation}
In the context of energy source representation, we shall refer to
distributions of point support, necessarily taking the form
\ref{eq:MPSappx}, as {\em multipole sources}, or simply {\em multipoles}.
Such sources combine
localization of energy and (possibly) anisotropic radiation pattern.
%distribution in space and time. ?? MJB
The coefficient distributions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
tensor-valued, according to the nature of the quantity (pressure, velocity, or stress) 
in the equation in which
$f(\mathbf x,t)$ appears as right-hand side.  Multipoles may
approximate arbitrary sources highly localized on the wavelength scale, in
the sense of generating approximately the same field away from the
source location, and for this reason have enjoyed widespread use in
the representation of seismic sources \cite[]{Shearer:2009}.

The isotropic acoustic point radiator problem \ref{eq:isorad} is
(equivalent to) a
simple special case of the symmetric hyperbolic system
\ref{eq:hyper_fam} with multipole
right hand sides $f,g$. Therefore gridded approximation of the more general
dynamics modeled by \ref{eq:hyper_fam} suffers from the difficulty
pointed out earlier for \ref{eq:isorad}.

Our results pertain to a specific class of finite difference
approximations for systems of the form \ref{eq:hyper_fam},
the so called {\em staggered grid schemes}, defined in detail in below
(equation \ref{eq:sg} and surrounding discussion). These schemes have the
virtue of obeying {\em energy estimates}, which are the chief
analytical tool in our work, and are widely used for basin-
and exploration-scale seismic modeling, acoustics computations,
structural dynamics, and many other purposes; see \cite{moczoetal:06} for an
excellent overview and many older references. Other classes of finite
difference approximations are also used in many of the same applications
\cite[]{Cohen:01,moczoetal:06,Petersson:2010,Petersson:2016}, as are finite element 
(spectral element, continuous and discontinous Galerkin) methods 
\cite[]{KomTromp:00,Cohen:01,Ghattas:IP25}. Results similar to those
reported here may hold for some of these approaches, however we do not
address this question here.

Staggered grid finite difference approximations $(u^n_h,v^{n+1/2}_h)$
to a solution  $(u,v)$ of
\ref{eq:hyper_fam} depend on the spatial grid cell size $h>0$ and the
time step $\Delta t>0$: $u^n_h$ approximates $u(\cdot,n\Delta t)$,
$v^{n+1/2}_h$ approximates $u(\cdot,(n+1/2)\Delta t)$. A sufficient
condition for consistent staggered grid schemes to produce convergent
approximations to 
smooth solutions of systems \ref{eq:hyper_fam} takes
the form of a {\em Courant-Friedrichs-Levy stability condition}, that is, a bound on
$\Delta t/h$, depending on the coefficient matrices $A,B$. In order to
express the error asymptotics of these schemes in a convenient way, we
will assume that $\Delta t$ is a function of $h$, obeying the
stability condition.

As pointed out earlier, point-wise convergence to singular solutions
cannot be expected, as the truncation error analysis on which it is
based makes no sense in the singular case. Instead, we turn to {\em
  weak} approximation: the approximate
solution $(u^n_h,v^{n+1/2}_h)$ converges weakly to continuum fields $(u,v)$ iff
arbitrary positively weighted averages of $(u^n_h,v^{n+1/2}_h)$
converge to the same weighted averages of
samplings of $(u,v)$ on the space-time $(h,\Delta t)$ grid, as $h,
\Delta t \rightarrow 0$ subject to the stability condition. 
The averaging weights are smooth functions of position (or position and time)
independent of grid, sampled at grid-points to create discrete weight functions.
Precise definitions are supplied below.

Our main result is that staggered grid approximations
$(u^n_h,v^{n+1/2}_h)$ 
converges weakly to a solution of \ref{eq:hyper_fam} with multipole right-hand side $(f,g)$, provided that the
finite-difference representation of the right-hand side
satisfies certain {\em discrete moment conditions}
\cite[]{Walden:1999,TorEng:04}, expressing accuracy of the 
right-hand side
approximation in a weak sense. ``Singular'' is used in
this paper as a synonym for {\em non-smooth distribution}, that is,
continuous linear functional on various topological vector spaces of
smooth functions, not representable by integration against a smooth kernel. For example, the right-hand side of the isotropic
radiator problem \ref{eq:isorad} defines a continuous linear
functional on the space of continuous functions with the sup norm. The
moment conditions used in this work simply specify that the numerical
approximations (linear combinations of grid values, with coefficients
depending on the grid diameter $h$) give the {\em same} result as the distribution being
approximated on polynomials with degree less than a bound depending on
the dimension $d$, the nature of the distribution (right-hand
side). This degree bound must be at least the {\em multipole order}
$N$, the highest derivative appearing in the definition
\ref{eq:MPSappx}. Express the degree bound as $N+q$, $q \ge 0$: we
call $q$ the {\em approximation order} of the discrete multipole representation.

The weak error asymptotics of a staggered grid scheme applied to a
problem with multipole right-hand side, using a discrete multipole
representation satisfying a discrete moment condition, depend on the
quantities introduced in the preceding paragraph, and on the error
asymptotics of the staggered grid scheme for smooth solutions. The
schemes we shall investigate produce approximations with error $O(\Delta t(h)^2 + h^p)$
(abbreviated as order $(2,p)$) to smooth solutions of systems \ref{eq:hyper_fam}.
%We give explicit rates of convergence, depending on on singular source
%approximation order $q$, spatial order $p$ for a staggered-grid
%finite difference scheme of with truncation error $O(\Delta t(h)^2,h^{p})$, the maximum multipole
%order between $f$ and $g$ denoted by $N^*$, and spatial dimension $d$.
%include explicitly the error bounds here
For spatial dimension $d$, multipole right-hand sides $(f,g)$  in
\ref{eq:hyper_fam}  of order $N$, discrete multipole
representation of approximation order $q$, and staggered grid scheme
of order $(2,p)$, we show that the error of averages with smooth averaging
functions $\tilde f, \tilde g$ satisfies the estimate
\begin{equation}
\label{eq:intro_weak_est}
\begin{split}
	\left| \int_0^T  \Big\{ \langle u,\tilde f\rangle + \langle v,\tilde g\rangle \Big\} dt -
		\Delta t \sum_{n=0}^N \Big\{ \langle u_h^{n}, \tilde f_h^{n} \rangle +
						 	   \langle v_h^{n+1/2}, \tilde g_h^{n+1/2}  \rangle \Big\} \right|\\
	= O(\Delta t(h)^2 + h^q + (\Delta t(h)^2+h^{p})h^{-N-d/2}).
\end{split}
\end{equation}
(Theorem \ref{thm:conv}). As discussed in the subsection on weak
convergence, staggered grid schemes typically involve multiple grids
with diameter $h$. In the estimate \ref{eq:intro_weak_est},
$\tilde{f}_h^{n}$ represents the samples on the grid(s) of diameter
$h$ of the smooth function $\bx \mapsto \tilde{f}(\bx,n\Delta t)$, and
similarly for $g$. Thus left-hand side of \ref{eq:intro_weak_est}
compares continuum ($L^2$) and discrete (weighted $l^2$) inner
products, the latter being the result of quadrature rules applied to
the former.

Note the implications of this estimate: to assure an error tending to
zero like $O(h^m)$, it is sufficient to require that $\Delta t(h) = O(h^{p/2})$, $q
\ge m$, $p \ge m +N+d/2$. These appear to be very stringent
requirements, compared to those for point-wise convergence to smooth solutions: for example for $N=0$ (that is, isotropic point source),
to achieve $m=2$ it is required that $q \ge 2$, $p \ge 2+d/2$ and $\Delta t(h) =
O(h^{1+d/4})$. In remarks after the proof of Theorem \ref{thm:conv},
we point out that if staggered grid approximations to {\em smooth} solutions can be shown to obey certain
$L^{\infty}$ estimates, as opposed to the $L^2$ estimates used here,
then the $h^{d/2}$ factor drops out of the right-hand side of the
estimate \ref{eq:intro_weak_est}, and the error asymptotics would
become optimal, that is, the same as those for smooth solutions. We
conjecture that such sup norm estimates for convergence of averages
hold, as our numerical results suggest, but their proof is beyond the
scope of this paper.

We assume throughout that the time-independent
coefficient matrices $A,B$ in the system \ref{eq:hyper_fam} are
smooth ($C^{\infty}$) in their dependence on the space variables. It
is certainly possible to extend some of these results to larger
classes of coefficients defined by weaker regularity
conditions - indeed, the weak solutions for systems of
the form \ref{eq:hyper_fam} exist if $A, B$ and their inverses are
uniformly bounded and measurable functions of the space variables
\cite[]{BlazekStolkSymes:13}. However the extent to which (and sense
in which) convergence
of finite difference methods might be established with such weak
regularity assumptions is an open question, beyond the scope of this work.

We analyze the approximation of waves in unbounded material models:
the spatial domain of wave propagation is the entire space $\bR^d$. We
consider only initial and right-hand side data of compact support in
space, so that the solutions of the wave propagation systems also have
compact support in space, for each time. Extension of our results to
settings in which waves have nontrivial interaction with boundaries is
possible in some cases, notably when the boundary conditions are
conservative or energy-dissipative. However we do not explicitly
formulate such generalizations, nor do we have anything to say about
sources located on physical boundaries. An effective extension to free
surface modeling in elasticity, for instance, is constrained by the
lack of conservative boundary conditions of order higher than
two. These are topics for further research.

We should note that numerical tests presented in the last section of
the paper suggest that, in the case of smooth coefficients, the actual 
quality of finite difference approximate solutions is better than our 
convergence result (or even the hypothetical ``optimal'') result
mentioned above) indicates: in some cases, away from the source 
point, we observe point-wise (strong) convergence with optimal 
order. We will mention this phenomenon again in the discussion 
section. The convergence behaviour suggested by the numerical examples 
is local, in regions bounded away from the source point. The 
convergence result proved here, on the other hand, is global.

>>>>

\title{Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems}
\date{}
\address{
        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
        77005-1892 USA
}
\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}

\righthead{Multipole Source Discretizations}

\maketitle
\parskip 12pt

@@

Seismic sources are commonly idealized as point-sources due
to their small spatial extent relative to seismic wavelengths.
The acoustic isotropic point-radiator is inadequate as a model 
of seismic wave generation for seismic sources that are known to exhibit directivity. 
Therefore, accurate modeling of seismic wavefields must include source 
representations generating anisotropic radiation patterns. 
Such seismic sources can be modeled as linear combinations of 
{\em multipole point-sources}.
In this paper we present a method for discretizing multipole sources 
in a finite difference setting, an extension of the moment matching 
conditions developed for the Dirac delta function in other applications.
We also provide the necessary analysis and numerical evidence 
to demonstrate the accuracy of our singular source approximations.
In particular, we develop a weak convergence theory for the discretization of 
a family of symmetric hyperbolic systems of first-order partial differential equations, with singular source terms, solved via staggered-grid finite difference methods.
Numerical experiments demonstrate a stronger result than 
what is presented in our convergence theory, namely, optimal convergence rates 
of numerical solutions are achieved point-wise in space away from the source if an 
appropriate source discretization is used.

@@

Seismic sources are commonly idealized as concentrated at a source point due
to their small spatial extent relative to seismic wavelengths.
A simple and familiar example of a wave propagation model with
spatially concentrated source is the isotropic point-radiator problem
for the acoustic wave equation in Euclidean 3-space \cite[]{CourHil:62}:
\begin{equation}\label{eq:isorad}
\begin{split}
        \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - c^2\nabla^2
  p(\mathbf x,t) & = f(\mathbf x,t) \equiv w(t)\delta(\mathbf x),\\ 
p(\mathbf x,t) & =  0,  \quad t<<0,
\end{split}
\end{equation}
$\delta(\mathbf x)$ being the Dirac delta function.
The solution $p$ is spherically symmetric:
\begin{equation}
\label{eq:green3d}
	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi c^2 r}, \quad 
	r= \sqrt{\mathbf x^T \mathbf x}.
\end{equation}
Both active and earthquake seismic sources, however, generate
spatially asymmetric wavefields (see for example \cite{Shearer:2009}, \cite{Yil:01}). 
The acoustic isotropic point-radiator 
is therefore inadequate as a model of seismic wave generation
and propagation because of its prediction of spatial symmetry.
The symmetry of the solution arises in part from the spherical symmetry of
the right-hand side in equation \ref{eq:isorad}. Therefore, accurate
modeling of seismic wavefields must include energy source (right-hand
side) representations generating anisotropic radiation patterns.

A {\em multipole}, or {\em multipole source} in the 
context of source modeling, is a finite linear combination of partial 
derivatives of the spatial Dirac delta function. Such sources combine
localization of energy and anisotropic radiation pattern. In fact,
Peetre's Theorem \cite[]{Horm:69} implies that any function of space
and time $f(\mathbf x,t)$ concentrated entirely at a point in space
(of {\em point support}) is a multipole of finite order $N \ge 0$,
\begin{equation}\label{eq:MPSappx}
        f(\mathbf x,t) = \sum_{|\mathbf s|=0}^N w_{\mathbf s}(t) \; D^{\mathbf s}\delta(\mathbf x-\mathbf x^*),
\end{equation}
in which we have introduced {\em multi-index notation}: for spatial
dimension $d=\{1,2,3\}$ and multi-index (integer d-tuple) ${\bf s} =
(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative operator, denoted $D^{\bf s}$,
and its (total) order $|{\bf s}|$, are defined as
\begin{equation}\label{eq:PDO}
        D^{\mathbf s} = \prod_{i=1}^d \left( \frac{\partial}{\partial x_i}\right)^{s_i}, \quad |\mathbf s| = \sum_{i=1}^{d} s_i.
\end{equation}
The coefficient time functions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
tensor-valued, according to the nature of the quantity (pressure, velocity, or stress) 
in the equation in which
$f(\mathbf x,t)$ appears as right-hand side.  Multipoles may
approximate arbitrary sources highly localized on the wavelength scale, in
the sense of generating approximately the same field away from the
source location, and for this reason have enjoyed widespread use in
the representation of seismic sources \cite[]{Shearer:2009}.

While finite element methods are also used in earthquake seismic modeling and inversion \cite[]{KomTromp:00,Cohen:01,Ghattas:IP25}, this paper focuses on regular (rectangular) grid finite difference methods, in particular the {\em staggered-grid} variety, which are widely used for basin- and exploration-scale modeling; see \cite{moczoetal:06} for an excellent overview and many older references. 
Such methods pose an immediate problem for singular source models such as multipoles: finite
difference algorithms ``know'' only gridded fields, so a source located at an arbitrary point $\mathbf x^*$ in space must be represented somehow by virtual sources at nearby grid points. 
This task is complicated by the nature of the field, as is evident for instance from inspection of the 3-D analytical function \ref{eq:green3d}: solutions of the acoustic equations with singular right-hand sides are generally themselves singular at the source point, so the Taylor-series based analysis of finite difference accuracy does not apply.
%Imitating finite-element singular source representation, for example by using adjoint interpolation, may reduce the accuracy of the modeled field, in the sense of convergence order: visually, large errors can pollute the field near the source, and propagate throughout the simulation domain.

The focus of this paper is to study singular source discretizations, particularly in the context of staggered-grid finite difference solvers for a family of wave propagation differential equations (e.g., linear acoustics and elasticity in first-order form) with multipole sources.
Specifically, we are interested in the following class of symmetric hyperbolic systems of first-order partial differential equations: find vector-valued fields $(u,v)$ such that
\begin{equation}\label{eq:hyper_fam}
\begin{split}
	A \frac{\partial}{\partial t}u + P^T v &= f \\
	B \frac{\partial}{\partial t}v  - P u & = g
\end{split}
\end{equation}
for given source terms $(f,g)$.
Operators $A$ and $B$ are bounded symmetric matrix-valued and represent medium parameters.
$P$ is a first-order spatial differential operator with constant coefficients, e.g., $P$ is the gradient in the acoustic equations \ref{eq:acoustics} with $P^T$ being the negative of the divergence.
We directly discretize multipole sources using {\em discrete moment conditions} as developed by \cite{Walden:1999}, and elaborated by \cite{TorEng:04} for multi-dimensional approximations.
Let $(u_h,v_h)$ denote the finite difference solution to \ref{eq:hyper_fam} via staggered-grid finite differences, with discretized source terms $(f_h,g_h)$, where $h>0$ denotes the grid size.
The class of staggered-grid finite difference methods we deal with are second-order in time and $2p$-order in space accurate, and we refer to such a scheme as being of order $2$-$2p$.

Our main contribution is a weak convergence theory, a novel approach to the singular source approximation problem, that is applicable to \ref{eq:hyper_fam} in that $(u_h,v_h)$ converge to $(u,v)$ given that source approximations $(f_h,g_h)$ converge to $(f,g)$ as $h\to 0$, both in a weak sense which we make clear in the theory section.
In particular, we show
\begin{equation}\label{eq:result}
	\langle (u,v),(\tilde f,\tilde g) \rangle - \langle (u_h,v_h), (\tilde f,\tilde g)\rangle = O(h^q + \Delta t^2 + (h^{2p}+\Delta t)h^{-N^*-d/2})
\end{equation}
given smooth test functions $(\tilde f,\tilde g)$. 
Inner products $\langle\cdot,\cdot\rangle$ are interpreted as standard $L^2$ space-time inner products in a continuum or discrete sense depending on their arguments.
Estimate \ref{eq:result} thus depends on singular source approximation order $q$, spatial half-order $p$ for a staggered-grid finite difference scheme of order $2$-$2p$, the maximum multipole order between $f$ and $g$ denoted by $N^*$, and spatial dimension $d$.
Numerical results presented here, consistent with other similar works (e.g., \cite{Petersson:2010}), appear to indicate, however, stronger convergence results: optimal convergence point-wise away from source.
In particular, we report second and fourth order rates when studying the spatial convergence of 2-2 and 2-4 finite difference methods respectively, where the source approximation order matched the spatial order of the numerical scheme.
